---
title: "project2"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment

The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.

## Download and load the data
```{r load}
library("data.table")
library("ggplot2")
library("dplyr")

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, destfile = 'repdata%2Fdata%2FStormData.csv.bz2')

stormDF <- read.csv("repdata%2Fdata%2FStormData.csv.bz2")
```


## Question 1
1. Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?

### Investigate which columns are necessary to answer the question

```{r subset_a}
health_damage <- stormDF %>% select(EVTYPE, FATALITIES, INJURIES)

#only keep data where fatalities and injuries are greater than 0
health_damage <- health_damage %>% filter(FATALITIES >0 | INJURIES >0)

#group by event type and get total for each damage
tot_healthDamage <- health_damage %>% group_by(EVTYPE) %>% 
        summarise(FATALITIES = sum(FATALITIES), 
                  INJURIES = sum(INJURIES))

# add in a total column that combines fatalities and injuries, and arrange by greatest total
tot_healthDamage <- tot_healthDamage %>% mutate(total = FATALITIES + INJURIES) %>% arrange(desc(total))

# only graph the top 10 events
top10_healthDamage<- tot_healthDamage[1:10,]

# melt the data frame to present it better in bar graph
top10_healthDamage <- melt(top10_healthDamage, id.vars="EVTYPE", variable.name = "type")


# create the bar graph
health_graph <- ggplot(top10_healthDamage, aes(x=reorder(EVTYPE, -value), y=value)) +
        geom_bar(stat="identity", aes(fill=type), position="dodge") +
        ylab("Frequency Count") +
        xlab("Event Type") +
        ggtitle("Top 10 Deadly Storm Events") + 
        theme(plot.title = element_text(hjust = 0.5))+
        theme(axis.text.x = element_text(angle=45, hjust=1))

health_graph
```


2. Across the United States, which types of events have the greatest economic consequences?
Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

```{r subset_b}
econ_damage <- stormDF %>% select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)

#filter to where damages > 0
econ_damage <- econ_damage %>% filter(PROPDMG> 0 | CROPDMG>0 )

#change all the values in PROPDMGEXP and  CROPDMGEXP to upper case:
econ_damage<- mutate_all(econ_damage, funs(toupper))

#convert data frame to data table
econ_damage <- as.data.table(econ_damage)

# map keys to values
propDmgKey <-  c("\"\"" = 10^0,
                 "-" = 10^0, 
                 "+" = 10^0,
                 "0" = 10^0,
                 "1" = 10^1,
                 "2" = 10^2,
                 "3" = 10^3,
                 "4" = 10^4,
                 "5" = 10^5,
                 "6" = 10^6,
                 "7" = 10^7,
                 "8" = 10^8,
                 "9" = 10^9,
                 "H" = 10^2,
                 "K" = 10^3,
                 "M" = 10^6,
                 "B" = 10^9)

# map keys to values
cropDmgKey <-  c("\"\"" = 10^0,
                 "?" = 10^0, 
                 "0" = 10^0,
                 "K" = 10^3,
                 "M" = 10^6,
                 "B" = 10^9)

econ_damage[, PROPDMGEXP := propDmgKey[as.character(econ_damage[,PROPDMGEXP])]]
econ_damage[is.na(PROPDMGEXP), PROPDMGEXP := 10^0 ]

econ_damage[, CROPDMGEXP := cropDmgKey[as.character(econ_damage[,CROPDMGEXP])] ]
econ_damage[is.na(CROPDMGEXP), CROPDMGEXP := 10^0 ]
        

#calculate damange cost for crop and property
econ_damage <- econ_damage %>% mutate(propDamCost = as.numeric(PROPDMG) *as.numeric(PROPDMGEXP)) %>%
        mutate(cropDamCost = as.numeric(CROPDMG)*as.numeric(CROPDMGEXP))

#group by event type and get total for each damage
tot_econ_damage <- econ_damage %>% group_by(EVTYPE) %>% 
        summarise(propDamCost = sum(propDamCost), 
                  cropDamCost = sum(cropDamCost))

# add in a total column and arrange by greatest total
tot_econ_damage <- tot_econ_damage %>% mutate(total = cropDamCost + propDamCost) %>% arrange(desc(total))

# only graph the top 10 events
top10_econDamage<- tot_econ_damage[1:10,]


# melt the data frame to present it better in bar graph
top10_econDamage <- melt(top10_econDamage, id.vars="EVTYPE", variable.name = "type")

# create the bar graph
econ_graph <- ggplot(top10_econDamage, aes(x=reorder(EVTYPE, -value), y=value)) +
        geom_bar(stat="identity", aes(fill=type), position="dodge") +
        ylab("Cost(US Dollars)") +
        xlab("Event Type") +
        ggtitle("Top 10 Economically Damaging Storm Events") + 
        theme(plot.title = element_text(hjust = 0.5))+
        theme(axis.text.x = element_text(angle=45, hjust=1))

econ_graph
```


